Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-03 analysis/paper/paper.qmd
Finished editing the draft!

A recent study conducted use-wear analysis on 95 stemmed points from the Jingeuneul site, along with examining fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were]. Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which likely occurred due to longitudinal forces from the shaft when a stone tip hits a target. They conclude that stemmed points may have been used as spear-throwers or bows [@sano2015backed].

@@

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site. For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones than in inland areas such as CT.

@@

ggplot(tcsa_mass_no_na,
       aes(`Weight(g)`,
           min_tcsa,
           colour = raw_materials_top)) +
  geom_point(size = 1.5) +
    labs(x = "Weight (g)", 
       y = "TCSA",
       colour='Raw material') +
  geom_smooth(method=lm, se=FALSE, size = .7)+
  theme_bw(base_size = 10) +
  theme(
  legend.title = element_text(size = 6.3),
  legend.text = element_text(size = 5.3),
  legend.position = c(.8,.3),
  legend.key.height = unit(4.8, 'mm'),
  legend.key.width = unit(2.5, 'mm'))

@@

# p-value
# simple version
# p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)

@@

 ggplot(tcsa_site,
              aes(y = min_tcsa, 
                  x = label_site)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "", 
       y = "TCSA") +
  theme(axis.text.x = element_text(size = 5)) +
  scale_x_discrete(guide = guide_axis(angle = 45))+
  geom_quasirandom(alpha = 0.4,
                   size = 1.6)+
   annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
     annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2)

@@

############ explore tcsa with MODERN eco_regions #######

library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))

veg_all <- st_transform(veg_all, crs = st_crs(4326))

# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()

pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)

sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762

points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)

eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

eco_box <- ggplot(points_in_veg_zones_all) +
  aes(ecoregion,
      min_tcsa) +
  geom_boxplot(size = 0.2) +
  geom_quasirandom(alpha = 0.3,
                   size = 1.3) +
  xlab("Vegetation") + 
  ylab("TCSA") +
  theme_minimal() +
    theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

# explore artifacts in certain zone, related to Tukey's test
tukey_veg <- points_in_veg_zones_all %>% 
  filter(ecoregion == "ST") %>% 
  dplyr::select(ecoregion, Assemblage, min_tcsa, raw_materials_top, age_ka)

############ explore tcsa with MODERN eco_regions & forest zones #######

veg_forest <- 
  read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))

veg_forest <- 
  st_transform(veg_forest, crs = st_crs(4326))

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()

pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)

#sf_use_s2(FALSE)

points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)

for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

# get rid of zones that has no artifact
points_in_veg_zones_forest_box <- 
  points_in_veg_zones_forest %>% 
  drop_na(Region)

for_box <- ggplot(points_in_veg_zones_forest_box) +
  aes(Zone,
      min_tcsa) +
  geom_boxplot(size = 0.2) +
  geom_quasirandom(alpha = 0.3,
                   size = 1.3) +
  xlab("Ecoregion") + 
  ylab("TCSA") +
  theme_minimal() +
      theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

######Tukey test######
#vegetation region#
points_in_veg_zones_all$ecoregion = as.factor(points_in_veg_zones_all$ecoregion)
eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)

eco_lgm_test <- Anova(lm(min_tcsa ~ ecoregion, points_in_veg_zones_all))

eco_lgm_test_pretty <- apastats::describe.Anova(eco_lgm_test, 1, 2)

eco_tur <- TukeyHSD(eco_aov)

# plot Tukey HSD
eco_tur_box <- 
eco_tur$ecoregion %>% 
  as_tibble(rownames = "zones") %>% 
  ggplot() +
  geom_linerange(aes(xmin = lwr, 
                     xmax = upr,
                     y = zones),
                 size = 0.4) +
  geom_point(aes(diff,
                 zones),
             size = 1) +
  geom_vline(xintercept = 0, colour = "red") +
  ylab("Pairwise comparison\nof TSCA in different\nvegetation zones") +
  xlab("Differences in means") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

#eco region#
for_aov <- 
  aov(min_tcsa ~ as.factor(Zone),
      points_in_veg_zones_forest_box)

# summary(for_aov)
for_lgm_test <- 
  Anova(lm(min_tcsa ~ as.factor(Zone),
           points_in_veg_zones_forest_box))

for_lgm_test_pretty <- 
  apastats::describe.Anova(for_lgm_test, 1, 2)

for_tur <- TukeyHSD(for_aov)

# plot Tukey HSD
for_tur_box <- 
for_tur$`as.factor(Zone)` %>% 
  as_tibble(rownames = "zones") %>% 
  mutate(zones = fct_reorder(zones, diff)) %>% 
  ggplot() +
  geom_linerange(aes(xmin = lwr, 
                     xmax = upr,
                     y = zones), 
                 size = 0.4) +
  geom_point(aes(diff,
                 zones),
             size = 1) +
  geom_vline(xintercept = 0, colour = "red") +
  ylab("Pairwise comparison\nof TSCA in different\necological zones") +
  xlab("Differences in means") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

@@

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits. TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B). We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

@@

Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones. Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second model in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).

@@

According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distribution of TCSA in each assemblage. Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic. From our findings, we infer that people were exposed to an unfamiliar situation and used a stemmed point as a multifunctional tool to carry out multiple tasks. We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we will conduct usewear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.

>>>>

@@

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site. For TCSA, they were only able to use 10 stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft. They conclude that stemmed points may have been used as spear-throwers or bows. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones and ESC of geographical zones than in inland areas such as CT of vegetation zones and CI of geographical zones.

@@

ggplot(tcsa_mass_no_na,
       aes(`Weight(g)`,
           min_tcsa,
           colour = raw_materials_top)) +
  geom_point(size = 1.5) +
    labs(x = "Weight (g)", 
       y = "TCSA",
       colour='Raw material') +
  geom_smooth(method=lm, se=FALSE, size = .7)+
  theme_bw(base_size = 10) +
  theme(
  legend.title = element_text(size = 6.3),
  legend.text = element_text(size = 5.3),
  legend.position = c(.8,.3),
  legend.key.height = unit(4.8, 'mm'),
  legend.key.width = unit(2.5, 'mm'))
```

```{r}
# p-value
# simple version
# p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)

@@

@@

 ggplot(tcsa_site,
              aes(y = min_tcsa, 
                  x = label_site)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "", 
       y = "TCSA") +
  theme(axis.text.x = element_text(size = 5)) +
  scale_x_discrete(guide = guide_axis(angle = 45))+
  geom_quasirandom(alpha = 0.4,
                   size = 1.6)+
   annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
     annotate("rect", 
           xmin = 7.5,
           xmax = 8,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2)
```

```{r}
######Tukey test######
#vegetation region#

library(sf)

############ explore tcsa with MODERN eco_regions #######
library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
veg_all <- st_transform(veg_all, crs = st_crs(4326))
# ggplot() +
#   geom_sf(data = veg_all) +
#   coord_sf()
pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_all))
# ggplot() +
#   geom_sf(data = pts_all)
sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
points_in_veg_zones_all <- 
st_join(veg_all, 
        pts_all, 
        join = st_contains)
eco_plot <- ggplot() +
  geom_sf(data = veg_all, aes(fill = "grey")) +
   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
  theme_minimal() +
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
  theme(axis.text.y = element_text(size = 4.5))+
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.position = "none")

eco_box <- ggplot(points_in_veg_zones_all) +
  aes(ecoregion,
      min_tcsa) +
  geom_boxplot(size = 0.2) +
  geom_quasirandom(alpha = 0.3,
                   size = 1.3) +
  xlab("Vegetation") + 
  ylab("TCSA") +
  theme_minimal() +
    theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6)) 
# explore artifacts in certain zone, related to Tukey's test
tukey_veg <- points_in_veg_zones_all %>% 
  filter(ecoregion == "ST") %>% 
  dplyr::select(ecoregion, Assemblage, min_tcsa, raw_materials_top, age_ka)

############ explore tcsa with MODERN eco_regions & forest zones #######
veg_forest <- 
  read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))
veg_forest <- 
  st_transform(veg_forest, crs = st_crs(4326))

# ggplot() +
#   geom_sf(data = veg_forest) +
#   coord_sf()
pts_forest <- 
  st_as_sf(korean_archaeological_site_locations_tcsa,
                coords = c("long_dd.y", "lat_dd.y"),
                crs = st_crs(veg_forest))
# ggplot() +
#   geom_sf(data = pts_forest)
#sf_use_s2(FALSE)
points_in_veg_zones_forest <- 
st_join(veg_forest, 
        pts_forest, 
        join = st_contains)
for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
  theme(legend.title = element_text(size= 6),
        legend.key.height = unit(4.5, 'mm'),
        legend.key.width = unit(2, 'mm'),
        legend.text = element_text(size = 4))

# get rid of zones that has no artifact
points_in_veg_zones_forest_box <- 
  points_in_veg_zones_forest %>% 
  drop_na(Region) 
for_box <- ggplot(points_in_veg_zones_forest_box) +
  aes(Zone,
      min_tcsa) +
  geom_boxplot(size = 0.2) +
  geom_quasirandom(alpha = 0.3,
                   size = 1.3) +
  xlab("Ecoregion") + 
  ylab("TCSA") +
  theme_minimal() +
      theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

######Tukey test######
#vegetation region#
points_in_veg_zones_all$ecoregion = as.factor(points_in_veg_zones_all$ecoregion)
eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)
eco_lgm_test <- Anova(lm(min_tcsa ~ ecoregion, points_in_veg_zones_all))
eco_lgm_test_pretty <- apastats::describe.Anova(eco_lgm_test, 1, 2)
eco_tur <- TukeyHSD(eco_aov)

# plot Tukey HSD
eco_tur_box <- 
eco_tur$ecoregion %>% 
  as_tibble(rownames = "zones") %>% 
  ggplot() +
  geom_linerange(aes(xmin = lwr, 
                     xmax = upr,
                     y = zones),
                 size = 0.4) +
  geom_point(aes(diff,
                 zones),
             size = 1) +
  geom_vline(xintercept = 0, colour = "red") +
  ylab("Pairwise comparison\nof TSCA in different\nvegetation zones") +
  xlab("Differences in means") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))
#eco region#
for_aov <- 
  aov(min_tcsa ~ as.factor(Zone),
      points_in_veg_zones_forest_box)
# summary(for_aov)
for_lgm_test <- 
  Anova(lm(min_tcsa ~ as.factor(Zone),
           points_in_veg_zones_forest_box))
for_lgm_test_pretty <- 
  apastats::describe.Anova(for_lgm_test, 1, 2)

for_tur <- TukeyHSD(for_aov)
# plot Tukey HSD
for_tur_box <- 
for_tur$`as.factor(Zone)` %>% 
  as_tibble(rownames = "zones") %>% 
  mutate(zones = fct_reorder(zones, diff)) %>% 
  ggplot() +
  geom_linerange(aes(xmin = lwr, 
                     xmax = upr,
                     y = zones), 
                 size = 0.4) +
  geom_point(aes(diff,
                 zones),
             size = 1) +
  geom_vline(xintercept = 0, colour = "red") +
  ylab("Pairwise comparison\nof TSCA in different\necological zones") +
  xlab("Differences in means") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(size = 4.5)) +
  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_text(size = 6),
          axis.title.y = element_text(size = 6))

@@

@@

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). @lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to data availability, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads. In general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).

The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, we found that variability decreased during the LGM in comparison to the previous period. This could be explained by the overall massive decline in stemmed points usage. We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites). However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability. Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

@@

Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones. Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).

@@

According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distribution of TCSA in each assemblage. Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic. From our findings, we infer that people were exposed to an unfamiliar situation and used a stemmed point as a multifunctional tool to carry out multiple tasks. We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
